---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "tools/readme/"
)
```

```{r echo=FALSE, message=FALSE}
library(devtools)
load_all()
set.seed(0)
```

# usedist

This package provides useful functions for distance matrix objects in R.

[![Travis-CI Build Status](https://travis-ci.org/kylebittinger/usedist.svg?branch=master)](https://travis-ci.org/kylebittinger/usedist)

## Installation

You can install usedist from github with:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("kylebittinger/usedist")
```

## Utility functions

Let's say we have a distance object representing the distances between six rows
of data:

```{r}
vals <- matrix(rnorm(30), nrow=6)
rownames(vals) <- c("A", "B", "C", "D", "E", "F")
vals
d <- dist(vals)
d
```

The `usedist` package allows us to select the distances for rows B, C, F, and
D, *in that order*:

```{r eval=FALSE}
library(usedist)
```


```{r}
dist_subset(d, c("B", "C", "F", "D"))
```

This is helpful when arranging a distance matrix to match a data frame, for 
instance with the `adonis` function in `vegan`.

We can extract distances between specified pairs of items, for example distances
between rows A-to-C and B-to-E. To do this, we provide two vectors of items: one
for the item of origin, and another for the destination.

```{r}
origin_row <- c("A", "B")
destination_row <- c("C", "E")
dist_get(d, origin_row, destination_row)
```

If items are arranged in groups, we can make a data frame listing the distances
between items in group 1 and group 2:

```{r}
item_groups <- rep(c("Control", "Treatment"), each=3)
dist_groups(d, item_groups)
```

Also, we provide a function to compute user-defined distances between
rows in the data matrix. For example, the Bray-Curtis dissimilarity is useful
for count data:

```{r}
cts <- matrix(rpois(30, 10), nrow=5)
rownames(cts) <- LETTERS[1:5]
cts
```

```{r}
bray_curtis_dissimilarity <- function (x1, x2) sum(abs(x1 - x2)) / sum(x1 + x2)
dist_make(cts, bray_curtis_dissimilarity)
```

## Centroid functions

The `usedist` package contains functions for computing the distance to group
centroid positions.  This is accomplished without finding the location of the
centroids themselves, though it is assumed that some high-dimensional Euclidean
space exists where the centroids can be situated.  References for the formulas
used can be found in the function documentation.

To illustrate, let's create a set of points in 2-dimensional space.  Four points
will be centered around the origin, and four around the point (3, 0).

```{r centroid_example}
pts <- data.frame(
  x = c(-1, 0, 0, 1, 2, 3, 3, 4),
  y = c(0, 1, -1, 0, 0, 1, -1, 0),
  Item = LETTERS[1:8],
  Group = rep(c("Control", "Treatment"), each=4))

library(ggplot2)
ggplot(pts, aes(x=x, y=y)) +
  geom_point(aes(color=Group)) +
  geom_text(aes(label=Item), hjust=1.5) +
  coord_equal()
```

Now for the trick: we'll take the distances between points, then figure out the
distances to the group centroids using only the point-to-point distances.

```{r}
pts_data <- pts[,1:2]
rownames(pts_data) <- pts$Item
pts_distances <- dist(pts_data)
pts_distances
```

First, we use the function `dist_between_centroids` to calculate the distance
between the centroids of the two groups.  In our example, we expect to get a
distance of 3.

```{r}
dist_between_centroids(
  pts_distances, c("A", "B", "C", "D"), c("E", "F", "G", "H"))
```

Now, we use the function `dist_to_centroids` to calculate the distance from each
individual point to the group centroids.  The within-group distances should all
be equal to 1.

```{r}
dist_to_centroids(pts_distances, pts$Group)
```

Double-checking the between-group distances is left as an exercise for the
reader.
